---
title: "R Notebook"
output: html_notebook
---

## TODO:
* write up an email, try to understand
* make a confusion matrix, at least for state to state
* make a chord diagram, using the census regions

```{r libraries}
library(tidyverse)
library(data.table)
# install.packages("devtools")
# devtools::install_github("UrbanInstitute/urbnmapr")
library(urbnmapr)
#install.packages("jpeg")
library(jpeg)
library(grid)
#install.packages("geosphere")
library(geosphere)
library(maps)
library(sf)
#install.packages("rgeos")
library(rgeos)
#install.packages("viridis")  # Install
library("viridis")  

```


# Visual Set 1: Movement in and out of the South

## Cleaning the Data
1. Load data
2. Remove margin of error columns
3. Convert the numerical columns from characters to numeric types (removing commas first)
4. Movement outside of the US is only specified at the country level and only includes inflows, not outflows. For simplification, the accompanying NAs in outflow and net migration will be converted to 0
5. Created a separate data table that only include movement that involves southern states

```{r cleaning}
movement_county_to_county <- fread("input/migration_county_to_county_2013_2017.csv")

# remove margin of error
movement_county_to_county <- select(movement_county_to_county, -ends_with("moe"))

# remove NAs (which are for rows that are not states)
movement_county_to_county <- filter(movement_county_to_county, is.na(a_state_code) == FALSE)

# convert numerical columns to numeric data type
cols_to_change <- 9:ncol(movement_county_to_county)
movement_county_to_county[cols_to_change] <-  as.numeric(str_remove(unlist(movement_county_to_county[cols_to_change]),","))
movement_county_to_county$a_state_name <- as.factor(movement_county_to_county$a_state_name)

# check NAs
lapply(movement_county_to_county, function(x) length(which(is.na(x) == TRUE)))

# convert NAs to 0
movement_county_to_county[cols_to_change] <- as.numeric(replace_na(unlist(movement_county_to_county[cols_to_change]),"0"))

# list of southern states
southern_states <- c(1, 5, 12, 13, 21, 22, 28, 37, 45, 47, 48, 51, 54)
names(southern_states) <- c("AL", "AR", "FL", "GA", "KY", "LA", "MS", "NC", "SC", "TN", "TX", "VA", "WV")    

# filter to include only movements involving southern states
south_county_to_county <- filter(movement_county_to_county, a_state_code %in% southern_states)

```

### All States
Created three summary tables:

| output table    | geography a  | geography b |
|-----------------|:------------:|:-----------:|
| county_to_state | county-level | state-level |
| state_to_state  | state-level  | state-level |
| state_all       | state-level  | -           |

Note: for state_all, inflows - outflows != net migration because we only have inflows from other countries, not outflows

```{r all}
county_to_state <- movement_county_to_county %>% 
  group_by(a_state_code, a_state_name, b_state_code, b_state_name, b_county_fips, b_county_name) %>%
  summarize(inflow = sum(inflow),
            outflow = sum(outflow), 
            net_migration = sum(net_migration))
fwrite(county_to_state, "output/county_to_state.csv")

state_to_state <- movement_county_to_county %>% 
  group_by(a_state_code, a_state_name, b_state_code, b_state_name) %>%
  summarize(inflow = sum(inflow),
            outflow = sum(outflow), 
            net_migration = sum(net_migration))
state_to_state <- setDT(state_to_state)
#removing migration within the state
state_to_state <- state_to_state[a_state_code != b_state_code]
fwrite(state_to_state, "output/state_to_state.csv")

state_all <- movement_county_to_county %>% 
  group_by(a_state_code, a_state_name) %>%
  summarize(inflow = sum(inflow),
            outflow = sum(outflow), 
            net_migration = sum(net_migration))
fwrite(state_all, "output/state_all.csv")
  
```


### Southern States

Filtered data to include only movement in and out of Southern states. 

For the purposes of this project, Southern states include: Alabama, Arkansas, Florida, Georgia, Kentucky, Louisiana, Mississippi, North Carolina, South Carolina, Tennessee, Texas, Virginia, and West Virginia

| output table           | geography a  | geography b |
|------------------------|:------------:|:-----------:|
| south_ccounty_to_state | county-level | state-level |
| south_cstate_to_state  | state-level  | state-level |
| south_cstate_all       | state-level  | -           |

```{r south}
south_county_to_state <- south_county_to_county %>% 
  group_by(a_state_code, a_state_name, b_state_code, b_state_name, b_county_fips, b_county_name) %>%
  summarize(inflow = sum(inflow),
            outflow = sum(outflow), 
            net_migration = sum(net_migration))
fwrite(south_county_to_state, "output/south_county_to_state.csv")

south_state_to_state <- south_county_to_county %>% 
  group_by(a_state_code, a_state_name, b_state_code, b_state_name) %>%
  summarize(inflow = sum(inflow),
            outflow = sum(outflow), 
            net_migration = sum(net_migration))
south_state_to_state <- setDT(south_state_to_state)
#removing migration within the state
south_state_to_state <- south_state_to_state[a_state_code != b_state_code]
fwrite(south_state_to_state, "output/south_state_to_state.csv")


south_state_all <- south_county_to_county %>% 
  group_by(a_state_code, a_state_name) %>%
  summarize(inflow = sum(inflow),
            outflow = sum(outflow), 
            net_migration = sum(net_migration))
fwrite(south_state_all, "output/south_state_all.csv")

```

## Basic Visuals

### Bar Graph: State Net Migration
```{r useful references, include=FALSE}
# https://www.datanovia.com/en/blog/ggplot-colors-best-tricks-you-will-love/#gradient-or-continuous-colors
# https://github.com/UrbanInstitute/urbnmapr
# http://www.endmemo.com/program/R/color.php 
```


```{r state bar graph}
# turns off scientific notation
options(scipen = 999)

state_all %>%
  arrange(desc(net_migration)) %>%
  ggplot(aes(x = reorder(a_state_name, net_migration), 
             y = net_migration, 
             fill = net_migration)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis(option = "plasma") +
  coord_flip() +
  theme_light() +
  ylab("Positive Net Migration, 2013-2017") +
  labs(title = "Net Migration") + 
  theme(legend.position = "none", 
        axis.title.y = element_blank()) +
  ggsave("state_to_state_bar.png", dpi = "retina")


```

### Bar Graph: Top County Movements

```{r top county movements}

# 8: Visualize the Biggest County Movements
summary_county %>% 
  arrange(desc(net_migration)) %>%
  top_n(50, net_migration) %>% 
  mutate(journey = paste(b_county_name, USPS.y, "to", a_county_name, USPS.x, sep = " ")) %>%
  ggplot(aes(x = reorder(journey, net_migration), 
             y = net_migration, 
             fill = net_migration)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis(option = "plasma") +
  coord_flip() +
  theme_light() +
  ylab("Positive Net Migration, 2013-2017") +
  labs(title = "Net Migration to the South: Top 50 Counties") +
  theme(legend.position = "none", 
        axis.title.y = element_blank()) +
  ggsave("southern_county_to_county_bar.png", dpi = "retina")
```

### Choropleth
```{r}
states_sf <- get_urbn_map(map = "states", sf = TRUE)

# join w/ movement data
states_all_sf <- full_join(states_sf, state_all, by = c("state_name" = "a_state_name"))

options(scipen = 999)
states_all_sf %>%
  ggplot() +
  geom_sf(aes(fill = net_migration),
          color = "snow", size = 0.5) +
  scale_fill_gradient2(midpoint = 0, low = "red", mid = "bisque1", high = "royalblue2", 
                       space = "Lab") + 
  labs(title = "Net Migration by State, 2013-2017", fill = "") +
  geom_sf_text(data = get_urbn_labels(map = "states", sf = TRUE), 
               aes(label = state_abbv), size = 3) +
  coord_sf(datum = NA) +
  ggsave("states_net_migration.png", device = "png")

```

## Complex Visuals

### Network Analysis with Gephi 
Nodes table:
* ID: generate unique
* Label: County, State Abrev (a and b combined into one list, i think only including the counties that end up in the analysis)
* State
* Region

```{r}
state_nodes <- data.table(ID = unique(state_to_state$b_state_code),
                          Label = unique(state_to_state$b_state_name))
fwrite(state_nodes, file = "state_nodes.csv")
```


Source: ** i think this will be only positive migration
* Source: ID of b state/county
* Target: ID of a state/county
* net_migration: to add weights
* inflow
* outflow

```{r}
state_edges <- state_to_state %>% transmute(Source = b_state_code,
                                            Target = a_state_code,
                                            inflow = inflow, 
                                            weight = inflow / 10000)
fwrite(state_edges, file = "state_edges.csv")
```

#### Network Analysis Visuals
State to state migration, representing the "inflow" amounts from state to state, rather than net migration. The network was filtered to include only inflows that were >= 10,000 and states that have 2 or more inflows of that size

*Version 1: Unweighted*:
INSERT IMAGE

*Version 2: Weighted*:
INSERT IMAGE

### State to State Migration Map

```{r my map}
# 2: load state latitude and longitude
state_latlong <- fread("input/statelatlong.csv")

# 3a: add lat/long for state a
summary_state <- left_join(state_to_state, state_latlong, by = c("a_state_name" = "state_name")) 
summary_state <- summary_state %>% rename(a_latitude = latitude)
summary_state <- summary_state %>% rename(a_longitude = longitude)

# 3b: add lat/long for state a
summary_state <- left_join(summary_state, state_latlong, by = c("b_state_name" = "state_name")) 
summary_state <- summary_state %>% rename(b_latitude = latitude)
summary_state <- summary_state %>% rename(b_longitude = longitude)

# 3c: remove rows where lon/lat are NA
summary_state <- na.omit(summary_state)

# 3d: remove alaska and hawaii, for simplicity
summary_state <- setDT(summary_state)
summary_state <- summary_state[a_state_name != "Alaska"]
summary_state <- summary_state[a_state_name != "Hawaii"]
summary_state <- summary_state[b_state_name != "Alaska"]
summary_state <- summary_state[b_state_name != "Hawaii"]

# 3e: remove negatives
summary_state <- summary_state[net_migration > 0]

# 4: rank by importance, aka by absolute value of net migration
#state_to_state <- state_to_state %>% mutate(positive = ifelse(net_migration >= 0, TRUE, FALSE),
#                              net_migration_abs = abs(net_migration))
summary_state <- summary_state %>% arrange(net_migration)



# 5: A function that makes a dateframe per connection (we will use these connections to plot each lines)
data_for_connection=function( dep_lon, dep_lat, arr_lon, arr_lat, group){
  inter <- gcIntermediate(c(dep_lon, dep_lat), 
                          c(arr_lon, arr_lat), 
                          n=50, 
                          addStartEnd=TRUE, 
                          breakAtDateLine=F)
  inter <- data.frame(inter)
  inter$group <- group
  return(inter)
}


# 6: create a combined data frame
data_plot <- data.frame()
for(i in c(1:nrow(summary_state))){
  tmp <- data_for_connection(summary_state$a_longitude[i], 
                             summary_state$a_latitude[i], 
                             summary_state$b_longitude[i], 
                             summary_state$b_latitude[i], 
                             i)
  tmp$net_migration <- summary_state$net_migration[i]
  data_plot <- rbind(data_plot, tmp)
}


# 7: Plot
ggplot() +
  geom_sf(data = st_as_sf(map("state", plot = FALSE, fill = TRUE)),  aes(), 
          fill = "black", color = "#f2f2f2", size = 1) +
  geom_sf_text(data = get_urbn_labels(map = "states", sf = TRUE), 
               aes(label = state_abbv), color = "white",
               size = 6) +
  geom_path(data = data_plot, 
            aes(x = lon, y = lat, group = group, color = log(net_migration), alpha = net_migration), 
            size = 2) +
  scale_color_viridis(option = "plasma") +
  #scale_fill_manual(values = c("bisque1", "#59a650")) +
  theme(
    #legend.position = "none",
    panel.background = element_rect(fill = "black", colour = "black"), 
    panel.spacing = unit(c(0,0,0,0), "null"),
    plot.margin = grid::unit(c(0,0,0,0), "cm"),
  ) +
  scale_x_continuous(expand = c(0.006, 0.006)) +
  ggsave("state_to_state_migration1.png", width = 26, height = 15.22, units = "in", dpi = "retina")

ggplot() +
  geom_sf(data = st_as_sf(map("state", plot = FALSE, fill = TRUE)),  aes(), 
          fill = "black", color = "#f2f2f2", size = 1) +
  geom_sf_text(data = get_urbn_labels(map = "states", sf = TRUE), 
               aes(label = state_abbv), color = "white",
               size = 6) +
  geom_path(data = data_plot, 
            aes(x = lon, y = lat, group = group, color = net_migration, alpha = net_migration), 
            size = 2) +
  scale_color_viridis(option = "plasma") +
  #scale_fill_manual(values = c("bisque1", "#59a650")) +
  theme(
    #legend.position = "none",
    panel.background = element_rect(fill = "black", colour = "black"), 
    panel.spacing = unit(c(0,0,0,0), "null"),
    plot.margin = grid::unit(c(0,0,0,0), "cm"),
  ) +
  scale_x_continuous(expand = c(0.006, 0.006)) +
  ggsave("state_to_state_migration2.png", width = 26, height = 15.22, units = "in", dpi = "retina")

```



### County to County Migration Map

This is a very large data set. To make it more intelligible visually, I made several adjustments: only included positive migration into Southern states, omitted movement between counties within the South

Even so, the resulting map had 32,466 connections. I then made two filtered maps - one with the top 10% of county migrations (~3,300) and one with the top 1% (~330)

```{r counties}
# 2a: load state latitude and longitude
# source: https://www.census.gov/geographies/reference-files/time-series/geo/gazetteer-files.html
county_latlong <- as.data.frame(fread("input/2019_Gaz_counties_national.csv", colClasses = "character"))
county_latlong <- county_latlong %>% mutate(INTPTLAT = as.numeric(INTPTLAT),
                                            INTPTLONG = as.numeric(INTPTLONG))
county_latlong$ANSICODE <- NULL
county_latlong$ALAND <- NULL
county_latlong$AWATER <- NULL
county_latlong$ALAND_SQMI <- NULL
county_latlong$AWATER_SQMI <- NULL

# 2b: impute combined state and county fips
south_county_to_county$a_fips <- as.character(paste(south_county_to_county$a_state_code,
                                                      formatC(south_county_to_county$a_county_fips,
                                                              width = 3, flag = "0"), sep = ""))
south_county_to_county$b_fips <- as.character(paste(south_county_to_county$b_state_code,
                                                      formatC(south_county_to_county$b_county_fips,
                                                              width = 3, flag = "0"), sep = ""))

# 3a - 1: add lat/long for state a --- all counties
#summary_county <- left_join(movement_county_to_county, county_latlong, by = c("a_fips" = "GEOID")) 
#summary_county <- summary_county %>% rename(a_latitude = INTPTLAT)
#summary_county <- summary_county %>% rename(a_longitude = INTPTLONG)

# 3a - 2: add lat/long for state a --- southern counties
summary_county <- left_join(south_county_to_county, county_latlong, by = c("a_fips" = "GEOID")) 
summary_county <- summary_county %>% rename(a_latitude = INTPTLAT)
summary_county <- summary_county %>% rename(a_longitude = INTPTLONG)

# 3b: add lat/long for state a
summary_county <- left_join(summary_county, county_latlong, by = c("b_fips" = "GEOID")) 
summary_county <- summary_county %>% rename(b_latitude = INTPTLAT)
summary_county <- summary_county %>% rename(b_longitude = INTPTLONG)

# 3c: remove rows where lon/lat are NA
summary_county <- na.omit(summary_county)

# 3d: remove alaska and hawaii, for simplicity
summary_county <- setDT(summary_county)
summary_county <- summary_county[a_state_name != "Alaska"]
summary_county <- summary_county[a_state_name != "Hawaii"]
summary_county <- summary_county[b_state_name != "Alaska"]
summary_county <- summary_county[b_state_name != "Hawaii"]

# 3e: remove movement within the same state
summary_county <- summary_county[a_state_name != b_state_name]

# 3f: remove movement between southern states
summary_county$b_state_code <- as.numeric(summary_county$b_state_code)
summary_county <- summary_county[b_state_code %in% southern_states == FALSE]

# 3g: remove negatives
summary_county <- summary_county[net_migration > 0]

# 4: visualize, to get a sense of distribution
summary_county %>% ggplot(aes(x = net_migration)) +
  geom_histogram()

table(summary_county$net_migration)

summary_county_top <- summary_county %>% top_frac(0.1, net_migration)

summary_county_top %>% ggplot(aes(x = net_migration)) +
  geom_histogram()

table(summary_county_top$net_migration)

# 4: rank by importance, aka by absolute value of net migration
# <- summary_county %>% mutate(positive = ifelse(net_migration >= 0, TRUE, FALSE),
#                              net_migration_abs = abs(net_migration))
summary_county <- summary_county %>% arrange(net_migration)

# 5: A function that makes a dateframe per connection (we will use these connections to plot each lines)
data_for_connection=function( dep_lon, dep_lat, arr_lon, arr_lat, group){
  inter <- gcIntermediate(c(dep_lon, dep_lat), 
                          c(arr_lon, arr_lat), 
                          n=50, 
                          addStartEnd=TRUE, 
                          breakAtDateLine=F)
  inter <- data.frame(inter)
  inter$group <- group
  return(inter)
}

# 6: create a combined data frame
data_plot_county2 <- data.frame()
for(i in c(1:nrow(summary_county))){
  tmp <- data_for_connection(summary_county$a_longitude[i], 
                             summary_county$a_latitude[i], 
                             summary_county$b_longitude[i], 
                             summary_county$b_latitude[i], 
                             i)
  tmp$net_migration <- summary_county$net_migration[i]
  data_plot_county2 <- rbind(data_plot_county2, tmp)
}

data_plot_county_top_10<- data.frame()
for(i in c(1:nrow(summary_county_top))){
  tmp <- data_for_connection(summary_county_top$a_longitude[i], 
                             summary_county_top$a_latitude[i], 
                             summary_county_top$b_longitude[i], 
                             summary_county_top$b_latitude[i], 
                             i)
  tmp$net_migration <- summary_county$net_migration[i]
  data_plot_county2 <- rbind(data_plot_county2, tmp)
}

# make a smaller dataset w/ top 10 largest in net migration
data_plot_county_top <- data_plot_county2 %>% top_frac(0.1, net_migration)



# 7a: Plot - all migration to southern counties
ggplot() +
  geom_sf(data = st_as_sf(map("state", plot = FALSE, fill = TRUE)),  aes(), 
          fill = "black", color = "#f2f2f2", size = 1) +
  geom_sf_text(data = get_urbn_labels(map = "states", sf = TRUE), 
               aes(label = state_abbv), color = "white",
               size = 6) +
  geom_path(data = data_plot_county2, 
            aes(x = lon, y = lat, group = group, color = net_migration, alpha = net_migration), 
            size = 2) +
  scale_color_viridis(option = "plasma") +
  #scale_fill_manual(values = c("bisque1", "#59a650")) +
  theme(
    #legend.position = "none",
    panel.background = element_rect(fill = "black", colour = "black"), 
    panel.spacing = unit(c(0,0,0,0), "null"),
    plot.margin = grid::unit(c(0,0,0,0), "cm")) +
  scale_x_continuous(expand = c(0.006, 0.006)) +
  ggsave("southern_county_to_county_migration.png", width = 26, height = 15.22, units = "in", dpi = "retina")


# 7b: Plot - top 10% of migration to southern counties
ggplot() +
  geom_sf(data = st_as_sf(map("state", plot = FALSE, fill = TRUE)),  aes(), 
          fill = "black", color = "#f2f2f2", size = 1) +
  geom_sf_text(data = get_urbn_labels(map = "states", sf = TRUE), 
               aes(label = state_abbv), color = "white",
               size = 6) +
  geom_path(data = data_plot_county2 %>% top_frac(0.10, net_migration), 
            aes(x = lon, y = lat, group = group, color = net_migration, alpha = net_migration), 
            size = 2) +
  scale_color_viridis(option = "plasma") +
  #scale_fill_manual(values = c("bisque1", "#59a650")) +
  theme(
    #legend.position = "none",
    panel.background = element_rect(fill = "black", colour = "black"), 
    panel.spacing = unit(c(0,0,0,0), "null"),
    plot.margin = grid::unit(c(0,0,0,0), "cm")) +
  scale_x_continuous(expand = c(0.006, 0.006)) +
  ggsave("southern_county_to_county_migration_top10.png", width = 26, height = 15.22, units = "in", dpi = "retina")


# 7c: Plot - top 1% of migration to southern counties
ggplot() +
  geom_sf(data = st_as_sf(map("state", plot = FALSE, fill = TRUE)),  aes(), 
          fill = "black", color = "#f2f2f2", size = 1) +
  geom_sf_text(data = get_urbn_labels(map = "states", sf = TRUE), 
               aes(label = state_abbv), color = "white",
               size = 6) +
  geom_path(data = data_plot_county2 %>% top_frac(0.01, net_migration), 
            aes(x = lon, y = lat, group = group, color = net_migration, alpha = net_migration), 
            size = 2) +
  scale_color_viridis(option = "plasma") +
  #scale_fill_manual(values = c("bisque1", "#59a650")) +
  theme(
    #legend.position = "none",
    panel.background = element_rect(fill = "black", colour = "black"), 
    panel.spacing = unit(c(0,0,0,0), "null"),
    plot.margin = grid::unit(c(0,0,0,0), "cm")) +
  scale_x_continuous(expand = c(0.006, 0.006)) +
  ggsave("southern_county_to_county_migration_top1.png", width = 26, height = 15.22, units = "in", dpi = "retina")

```


